\documentclass[12pt]{article}
\usepackage[english]{babel}
\usepackage{ucs}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath}
\usepackage{multirow}

\newcommand*\sepstars{%
  \begin{center}
    $\star\star\star$
  \end{center}
}

\title{Operations Research: Personal Notes}
\author{Paul Nechifor}

\begin{document}

\maketitle

\section{Terms}

\begin{description}

\item[algebraic method] \hfill
    \begin{itemize}
    \item can be used for any number of decision variables unlike the
        \emph{graphical method}
    \end{itemize}
    
    Say you have $n$ constraints and $m$ decision variables. You transform the
    inequalities from the constraints into equations by adding $n$ slack
    variables. Now you have $n+m$ total variables and a system of $n$ equations.
    
    You can find the corner points by setting $m$ of the variables to 0 and
    solving the system (some solutions will be unfeasable in which case they are
    not corner points). This means there are $C_{n+m}^m$ combinations (some
    unfeasable).
    
    (If you set the variables to any point other than 0 you
    might get inner points, not corner points.)

\item[basic solutions] \hfill \\
    The solutions you get by using the \emph{algebraic method}. If they are
    feasible they are called \emph{basic feasible solutions}.
    
    The variables which are set to 0 are the \emph{non-basic variables} and the
    ones which are solved are called \emph{basic variables}.

\item[constraints] \ldots

\item[corner points] \ldots

\item[decision variable] \ldots

\item[feasible region] \hfill \\
    All the feasible solutions. Is withing the corder points.

\item[feasible solution] \hfill \\
    Solutions which satisify all the constraints and the non-negativity
    requirements.

\item[graphical method] \hfill
    \begin{itemize}
    \item for two decision variables
    \end{itemize}
    
\item[non negativity restrictions] \ldots

\item[objective function] \ldots

\item[optimal solution] \hfill \\
    Something something.
    
\item[slack variables] \hfill \\
    The variables which are introduced to transform the inequalities in the
    constraints into equations. For example (1) into (2) by adding $x_3$.
    \begin{align}
        2x_1+5x_2&\ge3\\
        2x_1+5x_2+x_3&=3
    \end{align}
    They are subject to the non-negativity restrictions but do not count to the
    objective function.
    
    When $\ge$ inequalities are used, they are subtracted and are called
    \emph{negative slack variables} or \emph{surplus variables}. But still
    $x_i \ge 0$.

\end{description}

\section{Simplex method}

\subsection{Algebric form}

Example algorithm for:
\begin{align*}
    \text{max } 6x_1 + 5x_2 \\
    x_1 + x_2 &\le 5 \\
    3x_1 + 2x_2 &\le 12 \\
    x_1,x_2 &\ge 0
\end{align*}
It is transformed into:
\begin{align*}
    \text{max } 6x_1 + 5x_2 \\
    x_1 + x_2 + x_3 &= 5 \\
    3x_1 + 2x_2 + x_4 &= 12 \\
    x_1,x_2,x_3,x_4 &\ge 0
\end{align*}

The constraints should not have negative values so we write the equations by
$x_3$ and $x_4$ and set the rest to 0 (thus having a basic feasible solution).

Iteration 1:
\setcounter{equation}{0}
\begin{align}
    x_3 &= 5 - x_1 - x_2 \\
    x_4 &= 12 - 3x_1 - 2x_2 \\
    z &= 6x_1 + 5x_2 \notag
\end{align}

The objective is to increase $z$ so it appears that $x_1$ will increase it
faster. (1) limits $x_1$ to 5 and (2) limits it to 4. So we rewrite (2) by
$x_1$ and (1) with the new $x_1$. This is a new basic feasible solution.

Iteration 2:
\begin{align}
    3x_1 = 12 - 2x_2 - x_4 &\Leftrightarrow
        x_1 = 4 - \frac{2}{3}x_2 - \frac{1}{3}x_4 \\
    x_3 = 5 - (4 - \frac{2}{3}x_2 - \frac{1}{3}x_4) - x2 &\Leftrightarrow
        x_3 = 1 - \frac{1}{3}x_2 + \frac{1}{3}x_4 \\
    z = 6 (4 - \frac{2}{3}x_2 - \frac{1}{3}x_4) + 5x_2 &\Leftrightarrow
        z = 24 + x_2 - 2x_4 \notag
\end{align}

Now $z$ can be increased further by either incresing $x_2$ or decreasing $x_4$
(beacause of the negative coeficient $-2$). But decreasing $x_4$ would make it
negative so we have to increase $x_2$ which is limited to 6 in (3) and 3 in (4).

Iteration 3:
\begin{align}
    \frac{1}{3}x_2 = 1 - x_3 + \frac{1}{3}x_4 &\Leftrightarrow
        x_2 = 3 - 3x_3 + x_4 \\
    x_1 = 4 - \frac{2}{3}(3 - 3x_3 + x_4) &\Leftrightarrow
        x_1 = 2 + 2x_3 - x_4 \\
    z = 24 + (3 - 3x_3 + x_4) - 2x_4 &\Leftrightarrow
        z = 27 - 3x_3 - x_4 \notag
\end{align}

Since $x_3$ and $x_4$ cannot be increased, this is the final and optimal
solution.

Advantages of the simplex method:

\begin{itemize}
    \item Infeasiable solutions aren't evalutated.
    \item Evaluates progressively better solutions.
    \item Stops at the optimal solution.
\end{itemize}

\subsection{Tabular form}

Stages:

\begin{itemize}
    \item Initialization: Start with a basic feasible solution (the RHS
    variables are always non-negative).
    
    \item Iteration:
    
    \item Termination (when there is no entering column).
\end{itemize}

\begin{description}

\item[$c_j$] \hfill \\
    The coeficient of $x_j$ in the objective function.
    
\item[$\theta$] \hfill \\
    The limiting value for the chosen row. The one which limits it the most (the
    smallest) is choosen as the \emph{pivot row}.
    
    You don't compute $\theta$ for negative elements.
    
\item[entering column] \hfill \\
    \ldots
    
\item[pivot element] \hfill \\
    The intersection of the \emph{pivot row} and \emph{entering column}.
    
\item[pivot row] \hfill \\
    The leaving row.

\item[RHS] \hfill \\
    Right hand side.

\end{description}

\sepstars

\begin{itemize}
    \item The values for the row need to contain the identity matrix for the
    variables on the left (in that order).

    \item The \emph{non-basic variables} will have the values of $c_j - z_j$
    equal to 0.
\end{itemize}

\subsection{Minimization}

For this example:
\begin{align*}
    \text{min } 3x_1 + 4x_2 \\
    2x_1 + 3x_2 &\ge 8 \\
    5x_1 + 2x_2 &\ge 12 \\
    x_1,x_2 &\ge 0
\end{align*}
we transform it into
\begin{align*}
    \text{min } 3x_1 + 4x_2 + 0x_3 + 0x_4 \\
    2x_1 + 3x_2 - x_3 &= 8 \\
    5x_1 + 2x_2 - x_4 &= 12 \\
    x_1,x_2,x_3,x_4 &\ge 0
\end{align*}

Unlike in the first example, here $x_3$ and $x_4$ have negative coeficents. This
is done because $x_i \ge 0$ must hold for the minimum $\theta$ rule to work.

Simplex works for maximization problems. A minimization objective can be turned
into a maximization objective by negating the coeficents:

\begin{equation*}
    \text{max } -3x_1 - 4x_2 - 0x_3 - 0x_4
\end{equation*}

You can see that the slack variables do not qualify to be the initial basic
variables since they turn out negative. So the initial basic feasible solution
must be obtained by other means.

We rewrite it by adding two artificial variables with no meaning to the
problem: $a_1 = 8$ and $a_2 = 12$ (the corresponding RHS number):
\begin{align*}
    2x_1 + 3x_2 - x_3 + a_1 &= 8 \\
    5x_1 + 2x_2 - x_4 + a_2 &= 12
\end{align*}

\subsubsection{Big M method}

We have to modify the objective function and give $a_1$ and $a_2$ very small
contributions (thus preventing them from being choosen). So for maximization,
we give a very small number $-M$. It becomes:

\begin{equation*}
    \text{max } -3x_1 - 4x_2 - 0x_3 - 0x_4 - Ma_1 - Ma_2
\end{equation*}

$M$ is a large, positive, tends to infinity and is called big M.

$a_1$ and $a_2$ are choosen as the basic variables.

Artificial variables will never try to enter so there's no point in computing
$c_j-z_j$ for them.

\subsubsection{Two phase method}

This method eliminates big M which makes it better to be implemented on
computers.

The first phase eliminates the artificial variables and generates a basic
feasable solution.

The second phase generates the correct one.

With this method the objective function is changed so that all $x_i$ have a 0
contribution and the artificial variables have a $-1$ contribution (for a
maximization objective).

\end{document}
